Geometrization of characters of tori over non-Archimedean local fields

### 2013-08-12 shanghai.tex
Changes to the section on application to characters of tori.
I merged the two `applications' sections.

The category of geometrized characters of $G$, denoted by \cdef{$\GC(G)$},
is defined as follows.  A \cdef{geometrized character} on $G$ is a triple
$\cs{L}\ceq (\gcs{L},\mu,\phi)$ where:
\begin{enumerate}
\labitem{(CS.0)}{CS.0} $\gcs{L}$ is a constructible $\ell$-adic sheaf on  (the \'etale site of) $\bG$, locally constant of rank-$1$ on each connected component of $\bG$;
\labitem{(CS.1)}{CS.1} $\mu: \bm^* \gcs{L} \to \LxL$
  is an isomorphism of sheaves on $\bGxG$ such that the following diagram commutes, 
  where $m_3 \ceq m\circ (m\tight{1}{\times}{2}\id) = m\circ (\id\tight{2}{\times}{1} m)$;
%
  \[
  \begin{tikzcd}[row sep=30]
  \bm_3^*\gcs{L} \arrow{rr}{(\bm \tight{1}{\times}{2} \id)^*\mu} \arrow[swap]{d}{(\id \tight{2}{\times}{1} \bm)^*\mu}
    &&  \bm^*\gcs{L} \boxtimes \gcs{L} \dar{\mu \tight{0}{\boxtimes}{1} \id} \\
    \gcs{L} \boxtimes \bm^* \gcs{L} \arrow{rr}{\id \boxtimes \mu}
    &&  \gcs{L} \tight{0}{\boxtimes}{0} \LxL\\
  \end{tikzcd}
  \]

@@

\begin{proof}
As in Lemma~\ref{lemma:finite-pull-back}, we have $m^*(f,\psi) = (f_m,\psi_m)$ where $f_m$ and $\psi_m$ are pictured below, where $H_m = H\times_G(G\times G)$.
\[
\begin{tikzcd}
V & \Aut(V) & \arrow[swap]{l}{\psi} A \arrow[hook]{r} & H \arrow{r}{f} &  G\\
  &  & \arrow[dashed]{ul}{\psi_g} \arrow[swap]{u}{(m_f)\vert_{\ker m_f}} \ker f_m \arrow[hook]{r} & \arrow[swap]{u}{m_f} H_m \arrow{r}{f_m} & \arrow{u}{m} G\times G.
\end{tikzcd}
\]
On the other hand, set $p_1^*(f,\psi) = (f_1,\psi_1)$, so $f_1 = f\times \id : H\times G\to G\times G$ and $A_1\ceq \ker f_1 = A \times G$ and $\psi_1 = \psi \circ p_1$; set $p_2^*(f,\psi) = (f_2,\psi_2)$, so $f_2 = \id \times f : G\times H \to G\times G$ and $A_2\ceq \ker f_2 = G \times A$ and $\psi_2 = \psi \circ p_2$. Then $(f,\psi) \boxtimes (f,\psi) = (f_1,\psi_1) \otimes (f_2,\psi_2) = (f_1\times_{G\times G} f_2, \psi_1\otimes \psi_2) = (f\times f, \psi\otimes \psi) $ is given by
\[
\begin{tikzcd}
V\otimes V & \Aut(V\otimes V) & \arrow[swap]{l}{\psi\otimes \psi} A \times A \arrow[hook]{r}{} & H \times H \arrow{r}{f\times f} & G\times G\\
\end{tikzcd}
\]
Since $f\circ m_H = m \circ (f\times f)$, by the universal property of pull-back, there is a unique $\theta : H\times H \to H_m$ making the following commute.
\[
\begin{tikzcd}
\ &  & \arrow{ld}{\psi} A \arrow[hook]{r} & H \arrow{r}{f} &  G\\
V \arrow{d}{U} & \Aut(V)  & \arrow{l}{\psi_g} \arrow{u} A_m \arrow[hook]{r} & \arrow[swap]{u}{m_f} H_m \arrow{r}{f_m} & \arrow{u}{m} G\times G\\
V\otimes V & \Aut(V\otimes V) & \arrow[swap]{l}{\psi\otimes\psi} \arrow[bend left]{uu}{m_H} \arrow[swap]{u}{\theta\vert_{A\times A}} A \times A \arrow[hook]{r}{} & \arrow[bend left]{uu}{m_H} \arrow[swap]{u}{\theta} H \times H \arrow{r}{f\times f} & \arrow[equal]{u} G\times G\\
\end{tikzcd}
\]
Choose a basis $\{ v_1 \}$ for $V$; then $\{ v_1\otimes v_1 \}$ is a basis for $V\otimes V$. Define $U : V \to V\otimes V$ by $U(v_1) = v_1\otimes v_1$. Then the lower part of this diagram defines a weak isomorphism $(\theta,U): (f_m,\psi_m)\to (f\times f,\psi\otimes\psi)$. This defines an weak isomorphism $(\theta,U): m^*(f,\psi) \to (f,\psi)\boxtimes (f,\psi)$  in $C_1(G)$.

@@

We may now define a \cdef{global section}
of a geometrized character $\cs{L}$ on an \'etale group scheme $G$ as a choice $s(x) \in \gcs{L}_x$ for every
$x \in G(\bFq)$, satisfying
 \begin{equation}\label{section}
  \forall x,y \in G(\bFq), \qquad \mu_{x,y}(s(x+y)) = s(x)\otimes s(y).
 \end{equation}

@@

  It only remains to show that \eqref{pich1} is a group
  isomorphism.  Suppose that $\cs{L}$ and $\cs{L}'$ are two geometrized characters.
  If $s$ is a global section for $\cs{L}$ and $s'$ for $\cs{L}'$ then $s \otimes s'$ is a global section for
  $\cs{L} \otimes \cs{L}'$.  The three versions of \eqref{t} for $\cs{L}, \cs{L'}$ and $\cs{L} \otimes \cs{L}'$ imply that
  $\tau_{\cs{L} \otimes \cs{L}'}(\Frob{k}) = \tau_\cs{L}(\Frob{k}) \tau_{\cs{L}'}(\Frob{k})$.

  \Clifton{Straightforward, yes, but it still needs to be written down here.}
\end{proof}

@@

\section{Application to characters of bounded depth}

%\input{GdeN}
Let $p$ be a prime number. Let $K$ be a local field with residue field
$\Fq$, where $p$ divides $q$. Let $\OK$ be the ring of integers of $K$
and let $\pK$ be the maximal ideal in $\OK$.
%
In this section we introduce a group scheme $\GN{T}$ over $\Fq$
obtained by applying the Greenberg functor to the N\'eron lft-model
for $T$ and establish some basic facts about $\GN{T}$

@@

The lft-N\'eron model for $T$, denoted here by $\TT$, is a separated,
smooth, commutative group scheme locally of finite type over $\OK$
with generic fibre $T$, that enjoys the N\'eron mapping property: if
$X$ is a smooth scheme over $\OK$ then every morphism $X_\eta \to T$
over $K$ extends uniquely to a morphism $X \to \TT$. Equivalently, the
N\'eron mapping property states that the function
$\Hom_{\OK} (X,\TT) \to \Hom_K(X_\eta,T)$, given by restriction to
generic fibres, is a bijection; in particular, this implies $\TT(\OK) = T(K)$.

For each $d\in \ZZ_{>0}$, let $\TT_{d}$ be the smooth group scheme over
$\OK/\pK^d$ obtained from $\TT$ by base change along $\OK \to
\OK/\pK^d$. Observe that $\OK/\pK^d$ is an Artin ring and
$\TT_{d}$ is locally of finite type over $\OK/\pK^d$.

Let $\GN{T}_{d} = \Gr(\TT_d)$ be the group scheme over $\Fq$ produced by applying
the Greenberg functor to $\TT_{d}$. (See \cite{Greenberg:2} and
\cite[Ch. 9, \S 6]{BLR} for background on the Greenberg functor.)
%Then $\GN{T}_d(\Fq) = \TT(\OK/\pK^d)$.

\begin{proposition}\label{prop:GN}
  Fundamental properties of the group $\Fq$-scheme $\GN{T}_d$, for all $d\in \NN$:
  \begin{enumerate}
  \labitem{(GNd.0)}{GN0} $\GN{T}_d^0$, the identity component of
    $\GN{T}_d$, is a commutative algebraic group over $\Fq$;
  \labitem{(GNd.1)}{GNd} $\GN{T}_d$ is smooth and locally of finite type over $\Fq$;
  \labitem{(GNd.2)}{pi0} $\pi_0(\GN{T}_d) = \pi_0(\TT) \times_{\Spec{\OK}} \Spec{\Fq}$;
  \labitem{(GNd.3)}{fg} the component group of $\GN{T}_d$ is geometrically finitely generated;
  \labitem{(GNd.4)}{dFq} $\GN{T}_d(\Fq) = \TT(\OK/\pK^d)$, canonically;
\end{enumerate}
\end{proposition}

\begin{proof}
\Clifton{Proofs still missing, but all straightforward. I'll cite as much as I can, here.}
\end{proof}

%\input{application}
\begin{theorem}\label{thm:application}
  Let $T$ be an algebraic torus over a non-Archimedean local field $K$.
  Let $\GN{T}_d$ and $\GN{T}$ be as above.  For each $d \in \ZZ_{>0}$, the
  Picard group of the category of geometrized characters of $\GN{T}_d$ is
  canonically isomorphic to the group of characters of $T(K)$ of depth
  less than $d$:
  \[
  \GCiso{\GN{T}_d} \iso \Hom_{< d}(T(K),\EEx).
  \]
\end{theorem}

\begin{proof}
  By \ref{GNd} and \ref{fg}, we may apply Theorem~\ref{thm:snake} to
  $\GN{T}_d$ which provides a canonical isomorphism
  \[
  \GCiso{\GN{T}_d} \iso \Hom(\GN{T}_d(\Fq),\EEx).
  \]
  By \ref{dFq}, $\GN{T}_d(\Fq) \iso \TT(\OK/\pK^d)$ canonically.
  To finish the proof we recall that the filtration of $T(K)$ used to
  define depth has the property that $\TT(\OK/\pK^d) \cong T(K) / T(K)_d$
  \cite[Prop 5.2]{Yu} for $d \in \ZZ_{>0}$.  Thus characters of $\GN{T}_d(\Fq)$ are
  in bijection with characters of $T(K)$ vanishing on $T(K)_d$.
\end{proof}

\section{Application to admissible characters}

%\input{pro}
To conclude the paper we offer an interpretation of
Theorem~\ref{thm:application} that makes no mention of depth.

\Clifton{Explain what happens as $d$ varies.}
Suppose $d\leq d'$.

We may now assemble the group $\Fq$-schemes $\TT_{d}$ into one
glorious group pro-scheme over $\Fq$:
\begin{equation}\label{projlim}
  \GN{T} \ceq \varprojlim_{d\in \NN} \GN{T}_{d}.
\end{equation}
This is the \cdef{Greenberg transform of the lft-N\'eron model of $T$}; we will also write $\Gr$ for this functor on $\OK$-schemes.

\Clifton{Prove that $\GN{T}$ is a scheme? Strictly speaking, we don't
  care about \ref{scheme}.}

\Clifton{Must still define $\GC(\GN{T})$, since it is not an instance
 of our definitions, above.}

\begin{proposition}
  Fundamental properties of $\GN{T}$:
  \begin{enumerate}
  \labitem{(GN.0)}{scheme} $\GN{T}$ is a commutative group scheme over $\Fq$;
  \labitem{(GN.1)}{T(K)} $\GN{T}(\Fq) = T(K)$, canonically.
  \end{enumerate}
\end{proposition}

\begin{corollary}
  \[
  \GCiso{\GN{T}} \iso \Hom_\text{ad}(T(K),\EEx)
  \]
\end{corollary}

@@

>>>>

The category of geometrized characters of $G$, denoted by \cdef{$\GC(G)$},
is defined as follows.  A \cdef{geometrized character on $G$} is a triple
$\cs{L}\ceq (\gcs{L},\mu,\phi)$ where:
\begin{enumerate}
\labitem{(CS.0)}{CS.0} $\gcs{L}$ is a constructible $\ell$-adic sheaf on  (the \'etale site of) $\bG$, locally constant of rank-$1$ on each connected component of $\bG$;
\labitem{(CS.1)}{CS.1} $\mu: \bm^* \gcs{L} \to \LxL$
  is an isomorphism of sheaves on $\bGxG$ such that the following diagram commutes, 
  where $m_3 \ceq m\circ (m\tight{1}{\times}{2}\id) = m\circ (\id\tight{2}{\times}{1} m)$;
%
  \[
  \begin{tikzcd}[row sep=30]
  \bm_3^*\gcs{L} \arrow{rr}{(\bm \tight{1}{\times}{2} \id)^*\mu} \arrow[swap]{d}{(\id \tight{2}{\times}{1} \bm)^*\mu}
    &&  \bm^*\gcs{L} \boxtimes \gcs{L} \dar{\mu \tight{0}{\boxtimes}{1} \id} \\
    \gcs{L} \boxtimes \bm^* \gcs{L} \arrow{rr}{\id \boxtimes \mu}
    &&  \gcs{L} \tight{0}{\boxtimes}{0} \LxL\\
  \end{tikzcd}
  \]

@@

\begin{proof}
As in Lemma~\ref{lemma:finite-pull-back}, we have $m^*(f,\psi) = (f_m,\psi_m)$ where $f_m$ and $\psi_m$ are pictured below, where $H_m = H\times_G(G\times G)$.
\[
\begin{tikzcd}
V & \Aut(V) & \arrow[swap]{l}{\psi} A \arrow[hook]{r} & H \arrow{r}{f} &  G\\
  &  & \arrow[dashed]{ul}{\psi_g} \arrow[swap]{u}{(m_f)\vert_{\ker m_f}} \ker f_m \arrow[hook]{r} & \arrow[swap]{u}{m_f} H_m \arrow{r}{f_m} & \arrow{u}{m} G\times G.
\end{tikzcd}
\]
On the other hand, $(f,\psi) \boxtimes (f,\psi) = (f\times f, \psi\otimes \psi)$.
Since $f\circ m_H = m \circ (f\times f)$, by the universal property of pull-back, there is a unique $\theta : H\times H \to H_m$ making the following commute.
\[
\begin{tikzcd}
\ &  & \arrow{ld}{\psi} A \arrow[hook]{r} & H \arrow{r}{f} &  G\\
V \arrow{d}{U} & \Aut(V)  & \arrow{l}{\psi_g} \arrow{u} A_m \arrow[hook]{r} & \arrow[swap]{u}{m_f} H_m \arrow{r}{f_m} & \arrow{u}{m} G\times G\\
V\otimes V & \Aut(V\otimes V) & \arrow[swap]{l}{\psi\otimes\psi} \arrow[bend left]{uu}{m_H} \arrow[swap]{u}{\theta\vert_{A\times A}} A \times A \arrow[hook]{r}{} & \arrow[bend left]{uu}{m_H} \arrow[swap]{u}{\theta} H \times H \arrow{r}{f\times f} & \arrow[equal]{u} G\times G\\
\end{tikzcd}
\]
Choose a basis $\{ v_1 \}$ for $V$; then $\{ v_1\otimes v_1 \}$ is a basis for $V\otimes V$. Define $U : V \to V\otimes V$ by $U(v_1) = v_1\otimes v_1$. Then the lower part of this diagram defines a weak isomorphism $(\theta,U): m^*(f,\psi) \to (f,\psi)\boxtimes (f,\psi)$  in $C_1(G)$.

@@

We may now define a \cdef{global section}
of a geometrized character $\cs{L}$ on an \'etale group scheme $G$ as a choice $s(x) \in \gcs{L}_x$ for every
$x \in G(\bFq)$, satisfying
 \begin{equation}\label{section}
  \forall x,y \in G(\bFq), \qquad \mu_{x,y}(s(x+y)) = s(x)\otimes s(y).
 \end{equation}
\Clifton{I've been putting lemmata after the propositions that use them, 
to keep the flow of the sections clear. 
Would you mind if I do the same in this section?}

@@

  It only remains to show that \eqref{pich1} is a group
  isomorphism.  Suppose that $\cs{L}$ and $\cs{L}'$ are two geometrized characters.
  If $s$ is a global section for $\cs{L}$ and $s'$ for $\cs{L}'$ then $s \otimes s'$ is a global section for
  $\cs{L} \otimes \cs{L}'$.  The three versions of \eqref{t} for $\cs{L}, \cs{L'}$ and $\cs{L} \otimes \cs{L}'$ imply that
  $\tau_{\cs{L} \otimes \cs{L}'}(\Frob{k}) = \tau_\cs{L}(\Frob{k}) \tau_{\cs{L}'}(\Frob{k})$.
\end{proof}

@@

\section{Application to characters of algebraic tori over local fields}

Let $K$ be a local field with residue field $\Fq$.
%(Note that we place no restriction on the characteristic of $K$.)
Let $T$ be an algebraic torus over $K$.

%In this section we introduce a group scheme $\GN{T}$ over $\Fq$ obtained by applying the Greenberg functor to the N\'eron lft-model for $T$ and establish some basic facts about $\GN{T}$

@@

The lft-N\'eron model for $T$, denoted here by $\TT$, 
is a separated, smooth, commutative group scheme locally of finite type over $\OK$ 
(the ring of integers of $K$)
with generic fibre $T$, that enjoys the N\'eron mapping property: if
$S$ is a smooth scheme over $\OK$ then every morphism $S_\eta \to T$
over $K$ extends uniquely to a morphism $S \to \TT$. Equivalently, the
N\'eron mapping property states that the function
$\Hom_{\OK} (S,\TT) \to \Hom_K(S_\eta,T)$, given by restriction to
generic fibres, is a bijection; 
in particular, taking $S = \Spec{\OK}$, this implies $\TT(\OK) = T(K)$.

For each positive integer $d$, let $\TT_{d}$ be the group scheme obtained from $\TT$ by base change along $\OK \to \OK/\pK^d$, where $\pK$ is the maximal ideal in $\OK$. 
Observe that $\OK/\pK^d$ is an Artin ring and $\TT_{d}$ is smooth and locally of finite type over $\OK/\pK^d$.
%
Let \cdef{$\GN{T}_{d} = \Gr(\TT_d)$} be the group scheme over $\Fq$ produced by applying
the Greenberg functor to $\TT_{d}$; this is the \cdef{Greenberg transform of $\TT_d$}.
See \cite{Greenberg:2} and \cite[Ch. 9, \S 6]{BLR} and the first few sections of \cite{Stasinski} for background on the Greenberg functor. The fundamental properties of $\GN{T}_d$ that we need are:
 \begin{enumerate}
   \labitem{(GN.0)}{dFq} $\GN{T}_d(\Fq) = \TT(\OK/\pK^d)$, canonically.
  \labitem{(GN.1)}{GNd} $\GN{T}_d$ is a smooth, commutative group scheme, locally of finite type over $\Fq$, and the component group of $\GN{T}_d$ is geometrically finitely generated.
  \end{enumerate}
In fact, $\pi_0(\GN{T}_d) = \pi_0(\TT) \times_{\Spec{\OK}} \Spec{\Fq}$, and the latter is finitely generated by \cite{Xarles}.

%\input{application}
\begin{theorem}\label{thm:application}
  Let $T$ be an algebraic torus over a non-Archimedean local field $K$.
  Let $\GN{T}_d$ be as above.  For each positive integer $d$, the
  Picard group of the category of geometrized characters of $\GN{T}_d$ is
  canonically isomorphic to the group of characters of $T(K)$ of depth
  less than $d$:
  \[
  \GCiso{\GN{T}_d} \iso \Hom_{< d}(T(K),\EEx).
  \]
\end{theorem}

\begin{proof}
  By \ref{GNd}, we may apply Theorem~\ref{thm:snake} to
  $\GN{T}_d$ which provides a canonical isomorphism
  $\GCiso{\GN{T}_d} \iso \GN{T}_d(\Fq)^*$.
  By \ref{dFq}, $\GN{T}_d(\Fq) \iso \TT(\OK/\pK^d)$ canonically.
  To finish the proof we recall that the filtration of $T(K)$ used to
  define depth has the property that $\TT(\OK/\pK^d) \cong T(K) / T(K)_d$
  \cite[Prop 5.2]{Yu}, so characters of $\GN{T}_d(\Fq)$ 
  are in bijection with characters of $T(K)$ vanishing on $T(K)_d$. 
  %Note that $d$ is a positive integer.  
\end{proof}

%\section{Application to admissible characters}

%\input{pro}
We may also assemble the group $\Fq$-schemes $\TT_{d}$ into one
glorious group pro-scheme over $\Fq$:
\begin{equation}\label{projlim}
  \cdef{\GN{T}} \ceq \varprojlim_{d\in \ZZ_{>0}} \GN{T}_{d}.
\end{equation}
This is the \cdef{Greenberg transform of the N\'eron model of $T$}.
 Some fundamental properties of $\GN{T}$ are:
 \begin{enumerate}
   \labitem{(GN.2)}{T(K)} $\GN{T}(\Fq) = T(K)$, canonically;
  \labitem{(GN.3)}{scheme} $\GN{T}$ is a commutative group scheme over $\Fq$.
  \end{enumerate}
This gives us a natural way to geometrize all admissible characters of $T(K)$
without reference to depth.
Let \cdef{$\GC(\GN{T})$} be the ind-category $\varinjlim_{d\in \ZZ_{>0}} \GC(\GN{T}_d)$;
we refer to objects of $\GC(\GN{T})$ as \cdef{geometrized characters of $T$}.

\begin{corollary}
Let $T$ be an algebraic torus over a non-Archimedean local field $K$.
The Picard group of the category $\GC(\GN{T})$ of geometrized characters of $T$ is canonically isomorphic to the group of admissible characters of $T(K)$:
  \[
  \GCiso{\GN{T}} \iso \Hom_\text{ad}(T(K),\EEx).
  \]
\end{corollary}

@@

\bib{Stasinski}{article}{
   author={Stasinski, Alexander},
   title={Reductive group schemes, the Greenberg functor, and associated
   algebraic groups},
   journal={J. Pure Appl. Algebra},
   volume={216},
   date={2012},
   number={5},
   pages={1092--1101},
%   issn={0022-4049},
%   review={\MR{2875329}},
%   doi={10.1016/j.jpaa.2011.10.027},
}
